Weak gravitational lensing shear estimates with \textsc{metacalibration} for the \emph{Roman} High-Latitude Imaging Survey

### 2021-12-13 roman_metacal_v2.tex
v2.1 with simple simulations section added

Due to the subtlety in the lensing effect and its systematics-dominant nature in the weak regime, observational efforts have been challenging. Over the past decade, large international collaborations such as the Dark Energy Survey\footnote{\url{http://www.darkenergysurvey.org/}} (DES: \citealt{2005astro.ph.10346T}), the Hyper Supreme-Cam\footnote{\url{http://hsc.mtk.nao.ac.jp/ssp/}} (HSC: \citealt{2018PASJ...70S...4A}), and the Kilo-Degree Survey\footnote{\url{http://kids.strw.leidenuniv.nl/}} (KiDS: \citealt{2013ExA....35...25D}) have been successful at constraining the cosmological parameters, and the precision in calibrating the shapes of distant galaxies has reached a few percent (\citealt{2020arXiv201103408G, 2020arXiv201208567M, 2021A&A...645A.105G}). As we observe more area of the sky and develop better tools (observatories and algorithms), we detect more galaxies, and have already reached the point where statistical and systematic uncertainties are comparable. Thus, near-future Stage IV surveys (\citealt{2006astro.ph..9591A}) such as Euclid\footnote{\url{ http://sci.esa.int/euclid}} (Euclid Study Scientist \& the Science Advisory Team 2010), the Vera C. Rubin Observatory Legacy Survey of Space and Time\footnote{\url{ http://www.lsst.org}} (\emph{Rubin}: \citealt{2009arXiv0912.0201L, 2019ApJ...873..111I}), and \emph{Nancy Grace Roman} Space Telescope\footnote{\url{https://roman.gsfc.nasa.gov}} (\emph{Roman}: \citealt{2015arXiv150303757S}) require even better control of systematics. In order to understand the systematic uncertainties we face, we must develop realistic image simulations, apply pre-existing shape measurement to the simulated images, and determine any potential residual systematic effects and build a strategy for modifying the calibration methodology to mitigate them.

@@

Several shear calibration and bias mitigation efforts have been proposed by STEP (\citealt{2006MNRAS.368.1323H, 2007MNRAS.376...13M}) and GREAT (\citealt{2010MNRAS.405.2044B, 2013ApJS..205...12K, 2015MNRAS.450.2963M}) challenges, to meet the requirements for the current imaging surveys. In particular, one of the state-of-the-art self-calibration methods, \textsc{metacalibration} (\citealt{2017arXiv170202600H, 2017ApJ...841...24S}) has been shown to be able to substantially reduce the significance of shear bias. It has been confirmed that shear can be calibrated with \textsc{metacalibration} at the few-percent level in DES (\citealt{2018MNRAS.481.1149Z, 2020arXiv201103408G}), without accounting for galaxy blending and detection. Tackling the issue of blending/detection is one key step forward for the DES Y6 analysis and preparatory Rubin LSST work with the shear-dependent detection and calibration technique (\textsc{metadetect}; \citealt{2020ApJ...902..138S}). While the major advantage of \textsc{metacalibration} is that it can be directly applied to real galaxy images without needing to rely on an ensemble calibration from image simulations, limitations for future surveys are not well-known.

One possible limitation might lie in the effect of undersampled images in space-based misssions like \emph{Euclid} and \emph{Roman}, which have such a small pixel scale that images will not be Nyquist sampled in the near-infrared at the detector pixel scale (\citealt{2013PASP..125.1496S}) based on the Nyquist-Shannon Sampling Theorem. We define the sampling factor of each filter in \emph{Roman} to be, 
\begin{equation}
    Q = \frac{\lambda_{min}N_{f}}{p}, 
    \label{eqn:sampling}
\end{equation}
where $\lambda_{min}$ is the shortest wavelength of the incident light in the filter, $N_{f}$ is the focal ratio of the telescope ($N_{f}=7.8$ for \emph{Roman}) and $p$ is the pixel spacing of the sensor ($p=10\mu m$ for \emph{Roman}). An image with sampling factor $Q < 2$ is not Nyquist sampled (i.e., undersampled). The PSF needs to be interpolated and estimated from well-sampled images to deconvolve and measure shapes with, because the \emph{Roman} PSF has a complex structure that cannot be captured in the original undersampled images. It is, therefore, necessary to build a robust strategy to reconstruct well-sampled images to estimate the PSF to allow unbiased shape measurement.
Recently, \cite{2021MNRAS.502.4048K} (hereafter K21) addressed this potential issue of the limitations of \textsc{metacalibration} on undersampled images in \emph{Euclid} image simulations. They found that for the \emph{Euclid} mission the shear estimate is biased by about 1$\%$. It is mentioned that their result could be extended to the \emph{Roman} mission due to similarities in the instruments and for \emph{Roman} they predicted the multiplicative bias was more than 1\%. They show that these effects can be mitigated using additional weighting kernels in the measurement.

@@

%\subsection{Simple Simulations}
%\label{subsec:simplesim}
%The purpose of the simple simulations is to verify that there are no systematic biases at the basic levels of the simulation (e.g., no complex detector effects). The simple simulation is based on the real observational pointing strategy, and we use the real Sensor-Array Chips (SCAs) scheme, the band filters and the pixel scale as basic Roman specifications.

@@

\textsc{metacalibration} creates one unsheared and four artificially-sheared galaxy stamps. Each is deconvolved with the input PSF, left unsheared or sheared with an additional gravitational shear of $\delta\gamma=\pm 0.01$ in the two basis directions ($+g1$, $-g1$, $+g2$, $-g2$) and re-convolved with a new slightly larger isotropic, Gaussian PSF. \textsc{metacalibration} works under the assumption that the input shear is small, otherwise the recovered shear would be biased without accounting for higher order terms. It is a balance to choose an input shear that results in higher S/N of an object and unbiased shape measurement. We verify that our input gravitational shear is unbiased in \textsc{metacalibration} framework and the uncertainty in the recovered shape is small enough by constructing a simpler framework of the fiducial image simulations which renders Gaussian profile galaxies and PSFs on the \emph{Roman} pixel scale cutout image without any detector effects or other complications in the fiducial simulations. Figure \ref{fig:metacal_shear_linear} shows the performance as a function of input shear leading to the choice of input shear $\gamma=\pm 0.02$.

@@

For the single-band measurements, we can use the sampling factor defined in Equation \ref{eqn:sampling} per bandpass to show the relationship between the shear bias and image sampling. The sampling factor is an indication of how undersampled the image in some bandpass is. The sampling factor for each single-band measurement can be found in Table \ref{tab:bias_summary}. This relationship between sampling factor and shear bias was previously explored in \citealt{2021MNRAS.502.4048K} using \emph{Euclid} simulation with different shape measurement algorithms, where they found that the estimated multiplicative bias has a relatively large dependence on the sampling factor. Their result can be extrapolated to our single-band single-epoch results. The multiplicative bias of single-band single-epoch measurements is roughly 1\% in each filter. When the coadded images for each filter are used, the bias is at the similar level or worse within the uncertainties. We can conclude that in the current \emph{Roman} simulation the multiplicative bias has a similar dependence on the sampling factor as shown in K21, but with a smaller amplitude. We do not see any clear improvements in the recovered shear when we coadd the single-epoch images, despite having a smaller pixel scale (better sampling) in the images. This is most likely due to an issue with how we are currently treating the PSF or correlated noise in the coadding process, given the large observed $c_1$ when coadding discussed below.

@@

\begin{table*}
	\centering
	\label{tab:bias_summary}
	\begin{tabular}[scale=1.4]{ c|c|c|c|c|c } 
		\hline
		simulation variants & sampling factor ($Q$) & $m_{1}\times10^{-2}$ & $m_{2}\times10^{-2}$ & $c_{1}\times10^{-4}$ & $c_{2}\times10^{-4}$\\
		\hline
		single-epochs (J129) & 1.01 & 1.17$\pm$0.68 & 0.44$\pm$0.75 & 4.46$\pm$1.32 & 0.64$\pm$1.31\\
		single-epochs (H158) & 1.23 & -1.07$\pm$0.62 & -1.61$\pm$0.66 & 2.83$\pm$1.35 & 2.33$\pm$1.22\\
		single-epochs (F184) & 1.44 & -0.86$\pm$0.72 & -0.69$\pm$0.83 & 2.62$\pm$1.56 & 0.23$\pm$1.53\\
		\hline
		single-band coadd (J129) & 1.26 & -1.84$\pm$0.71 & -1.90$\pm$0.65 & 11.99$\pm$1.39 & 0.59$\pm$1.41\\
		single-band coadd (H158) & 1.54 & -1.91$\pm$0.67 & -2.11$\pm$0.67 & 13.99$\pm$1.34 & 5.40$\pm$1.35\\
		single-band coadd (F184) & 1.79 & -0.28$\pm$0.83 & -0.27$\pm$0.78 & 12.15$\pm$1.56 & 5.66$\pm$1.56\\
		\hline
		multiband single-epochs & N/A & -1.07$\pm$0.66 & -1.61$\pm$0.67 & 2.83$\pm$1.30 & 2.33$\pm$1.22 \\
		multiband coadd & N/A & -2.32$\pm$0.64 & -2.34$\pm$0.63 & 14.45$\pm$1.10 & 4.49$\pm$1.22\\

@@

\begin{figure*}
    \hspace*{-3.5cm}
    \centering
	\includegraphics[scale=0.34]{H158_meanshear_measured_properties_perbin_e1_v2.pdf}
    \caption{(Working on fixing the figure bounds)This shows the mean shear residual ($\Delta e_{1}$; $\langle e_{1,measured} \rangle - \langle e_{1,expected} \rangle$) as a function of various properties of galaxies for H158 single-band single-epochs, coadds, and multiband measurements. In this plot, the shear response is computed in each bin. The properties include the S/N, measured galaxy size $T_{gal,measured}$, input galaxy size $T_{gal,input}$, the shape and size of the input PSF.}
    \label{fig:meanshear}
\end{figure*}

>>>>

Due to the subtlety in the lensing effect and its systematics-dominant nature in the weak regime, observational efforts have been challenging. Over the past decade, large international collaborations such as the Dark Energy Survey\footnote{\url{http://www.darkenergysurvey.org/}} (DES: \citealt{2005astro.ph.10346T}), the Hyper Supreme-Cam\footnote{\url{http://hsc.mtk.nao.ac.jp/ssp/}} (HSC: \citealt{2018PASJ...70S...4A}), and the Kilo-Degree Survey\footnote{\url{http://kids.strw.leidenuniv.nl/}} (KiDS: \citealt{2013ExA....35...25D}) have been successful at constraining the cosmological parameters, and the precision in calibrating the shapes of distant galaxies has reached a few percent (\citealt{2020arXiv201103408G, 2020arXiv201208567M, 2021A&A...645A.105G}). As we observe more area of the sky and develop better tools (observatories and algorithms), we detect more galaxies, and have already reached the point where statistical and systematic uncertainties are comparable. Thus, near-future Stage IV surveys (\citealt{2006astro.ph..9591A}) such as Euclid\footnote{\url{ http://sci.esa.int/euclid}} (\citealt{2011arXiv1110.3193L}), the Vera C. Rubin Observatory Legacy Survey of Space and Time\footnote{\url{ http://www.lsst.org}} (\emph{Rubin}: \citealt{2009arXiv0912.0201L, 2019ApJ...873..111I}), and \emph{Nancy Grace Roman} Space Telescope\footnote{\url{https://roman.gsfc.nasa.gov}} (\emph{Roman}: \citealt{2015arXiv150303757S}) require even better control of systematics. In order to understand the systematic uncertainties we face, we must develop realistic image simulations, apply pre-existing shape measurement to the simulated images, and determine any potential residual systematic effects and build a strategy for modifying the calibration methodology to mitigate them.

@@

Several shear calibration and bias mitigation efforts have been proposed by STEP (\citealt{2006MNRAS.368.1323H, 2007MNRAS.376...13M}) and GREAT (\citealt{2010MNRAS.405.2044B, 2013ApJS..205...12K, 2015MNRAS.450.2963M}) challenges, to meet the requirements for the current imaging surveys. In particular, one of the state-of-the-art self-calibration methods, \textsc{metacalibration} (\citealt{2017arXiv170202600H, 2017ApJ...841...24S}) has been shown to be able to substantially reduce the significance of shear bias. It has been confirmed that shear can be calibrated with \textsc{metacalibration} at the few-percent level in DES (\citealt{2018MNRAS.481.1149Z, 2020arXiv201103408G}), without accounting for galaxy blending and detection. Tackling the issue of blending/detection is one key step forward for the DES Y6 analysis and preparatory Rubin LSST work with the shear-dependent detection and calibration technique (\textsc{metadetection}; \citealt{2020ApJ...902..138S}). While the major advantage of \textsc{metacalibration} is that it can be directly applied to real galaxy images without needing to rely on an ensemble calibration from image simulations, limitations for future surveys are not well-known.

One possible limitation might lie in the effect of undersampled images in space-based misssions like \emph{Euclid} and \emph{Roman}, which have such a small pixel scale that images will not be Nyquist sampled in the near-infrared (NIR) at the detector pixel scale (\citealt{2013PASP..125.1496S}) based on the Nyquist-Shannon Sampling Theorem. We define the sampling factor of each filter in \emph{Roman} to be, 
\begin{equation}
    Q = \frac{\lambda_{min}N_{f}}{p}, 
    \label{eqn:sampling}
\end{equation}
where $\lambda_{min}$ is the shortest wavelength of the incident light in the filter, $N_{f}$ is the focal ratio of the telescope ($N_{f}=7.8$ for \emph{Roman}) and $p$ is the pixel spacing of the sensor ($p=10\mu m$ for \emph{Roman}). An image with sampling factor $Q < 2$ is not Nyquist sampled (i.e., undersampled). The PSF needs to be interpolated and estimated from well-sampled images to deconvolve and measure shapes with, because the \emph{Roman} PSF has a complex structure that cannot be captured in the original undersampled images. It is, therefore, necessary to build a robust strategy to reconstruct well-sampled images to estimate the PSF to allow unbiased shape measurement.
Recently, \cite{2021MNRAS.502.4048K} (hereafter K21) addressed this potential issue of the limitations of \textsc{metacalibration} on undersampled images in \emph{Euclid} image simulations. They found that for the \emph{Euclid} mission the shear estimate is biased by about 1$\%$. It is mentioned that their result could be extended to the \emph{Roman} mission due to similarities in the instruments and for \emph{Roman} they predicted the multiplicative bias was more than 1\%. They show that these effects can be mitigated using additional weighting kernels in the measurement.

@@

\subsection{Validations with Simple Simulations}
\label{subsec:simplesim}
The purpose of the simple simulations is to verify that there are no systematic biases at the fundamental levels of the simulation (e.g., no complex detector effects). These simple simulations play roles in ruling out the sources of systematic biases that may potentially be a problem in the use of \textsc{metacalibration} with space-based images.

The parameter choices are listed in Table. \ref{tab:params}. With these parameters set, we simulated images with varying galaxy model profile, PSF model profile, the artificial shear, and background noise level, without any complications in the images and pipelines such as detector effects and coaddition. We first verify that our input gravitational shear is unbiased in \textsc{metacalibration} framework and the uncertainty in the recovered shape is small enough for the \emph{Roman} requirement (Fig. \ref{fig:metacal_shear_linear}). We then test the pipeline by doubling the background noise. Due to the \emph{Roman} being in NIR range, the expected background noise level in the images is higher compared to optical imaging surveys like DES. With this test, we can confirm that the treatment of correlated noise induced in \textsc{metacalibration} process for low S/N objects does not trigger any bias at the level we should be concerned (\citealt{2017ApJ...841...24S}). Another test with \emph{Roman} PSF instead of Gaussian PSF does guarantee that the PSF features particular to \emph{Roman} do not bias the shape recovery even though it is undersampled. Upon validating there does not exist any bias at the fundamental level, we began constructing coadd and more complex measurement pipeline. 
\begin{table}
    \centering
    \begin{tabular}[width=\columnwidth]{|p{3cm}||p{3cm}|p{3cm}|p{3cm}|}
    \hline
    Parameter & Value \\
    \hline
    Pixel scale & 0.11 arcsec/pixel\\
    Dither, SCA & 22535, 1 \\ 
    Bandpass & H158 \\
    Half-light radius & 1.0 arcsec\\
    Magnitude & randomly chosen from \emph{CANDELS} \\ 
    Stamp size & 32 pixels each side\\
    \hline
    \end{tabular}
    \caption{List of the parameter choices for the simple simulations.}
    \label{tab:params}
\end{table}

\begin{table*}
	\centering
	\label{tab:simple_sim}
	\begin{tabular}[width=\textwidth]{ c|c|c|c|c|c|c|c|c|c } 
		\hline
		simulation variants & galaxy profile & PSF profile & shear & background noise level & $m_{1}\times10^{-2}$ & $m_{2}\times10^{-2}$ & $c_{1}\times10^{-4}$ & $c_{2}\times10^{-4}$\\
		\hline
		Basic-0.02 & Gaussian & Gaussian & 0.02 & roman.getSkyLevel & 0.01$\pm$0.10 & -0.02$\pm$0.10 & -0.02$\pm$0.14 & 1.06$\pm$0.14\\
		Basic-0.05 & Gaussian & Gaussian & 0.05 & roman.getSkyLevel & 0.23$\pm$0.07 & 0.22$\pm$0.07 & 0.05$\pm$0.33 & 1.08$\pm$0.33\\
		Basic-0.1 & Gaussian & Gaussian & 0.1 & roman.getSkyLevel & 0.99$\pm$0.03 & 0.99$\pm$0.03 & 0.13$\pm$0.33 & 0.88$\pm$0.33\\
		\hline
		Doublesky-0.02 & Gaussian & Gaussian & 0.02 & 2 $\times$ roman.getSkyLevel & -0.12$\pm$0.18 & 0.01$\pm$0.18 & 0.06$\pm$0.36 & 1.04$\pm$0.36\\
		\hline
		RomanPSF-0.02 (non-rotated) & Gaussian & Roman & 0.02 & roman.getSkyLevel & 0.13$\pm$0.10 & 0.18$\pm$0.10 & -0.02$\pm$0.19 & 5.33$\pm$0.19\\
		\hline
	\end{tabular}
	\caption{This table compares the shear calibration bias, both multiplicative and additive bias for different simulation runs.}
	\label{tab:result}
\end{table*}

@@

\textsc{metacalibration} creates one unsheared and four artificially-sheared galaxy stamps. Each is deconvolved with the input PSF, left unsheared or sheared with an additional gravitational shear of $\delta\gamma=\pm 0.01$ in the two basis directions ($+g1$, $-g1$, $+g2$, $-g2$) and re-convolved with a new slightly larger isotropic, Gaussian PSF. \textsc{metacalibration} works under the assumption that the input shear is small, otherwise the recovered shear would be biased without accounting for higher order terms. It is a balance to choose an input shear that results in higher S/N of an object and unbiased shape measurement. Figure \ref{fig:metacal_shear_linear} shows the performance as a function of input shear leading to the choice of input shear $\gamma=\pm 0.02$.

@@

For the single-band measurements, we can use the sampling factor defined in Equation \ref{eqn:sampling} per bandpass to show the relationship between the shear bias and image sampling. The sampling factor is an indication of how undersampled the image in some bandpass is. The sampling factor for each single-band measurement can be found in Table \ref{tab:bias_summary}. This relationship between sampling factor and shear bias was previously explored in \citealt{2021MNRAS.502.4048K} using \emph{Euclid} simulation with different shape measurement algorithms, where they found that the estimated multiplicative bias has a relatively large dependence on the sampling factor. Their result can be extrapolated to our single-band single-epoch results. The multiplicative bias of single-band single-epoch measurements is roughly 1\% in each filter. When the coadded images for each filter are used, the bias is at the similar level or worse within the uncertainties. We can conclude that in the current \emph{Roman} simulation the multiplicative bias has a similar dependence on the sampling factor as shown in K21, but with a smaller amplitude. We do not see any clear improvements in the recovered shear when we coadd the single-epoch images, despite having a smaller pixel scale (better sampling) in the images. This is most likely due to an issue with how we are currently treating the PSF or correlated noise in the coadding process, given the large observed $c_1$ when coadding discussed below. \textcolor{red}{Although we have verified that the handling of correlated noise in \textsc{metacalibration} proess for single-epoch images is unbiased, it is possible that the \texttt{psc} process induces more correlated noise that is not corrected further. }

@@

\begin{table*}
	\centering
	\label{tab:bias_summary}
	\begin{tabular}[width=\textwidth]{ c|c|c|c|c|c } 
		\hline
		simulation variants & sampling factor ($Q$) & $m_{1}\times10^{-2}$ & $m_{2}\times10^{-2}$ & $c_{1}\times10^{-4}$ & $c_{2}\times10^{-4}$\\
		\hline
		single-epochs (J129) & 1.01 & 1.17$\pm$0.68 & 0.44$\pm$0.75 & 4.46$\pm$1.32 & 0.64$\pm$1.31\\
		single-epochs (H158) & 1.23 & -1.07$\pm$0.62 & -1.61$\pm$0.66 & 2.83$\pm$1.35 & 2.33$\pm$1.22\\
		single-epochs (F184) & 1.44 & -0.86$\pm$0.72 & -0.69$\pm$0.83 & 2.62$\pm$1.56 & 0.23$\pm$1.53\\
		\hline
		single-band coadd (J129) & 1.26 & -1.84$\pm$0.71 & -1.90$\pm$0.65 & 11.99$\pm$1.39 & 0.59$\pm$1.41\\
		single-band coadd (H158) & 1.54 & -1.75$\pm$0.67 & -1.68$\pm$0.64 & 14.39$\pm$1.32 & 4.29$\pm$1.35\\
		single-band coadd (F184) & 1.79 & -0.28$\pm$0.83 & -0.27$\pm$0.78 & 12.15$\pm$1.56 & 5.66$\pm$1.56\\
		\hline
		multiband single-epochs & N/A & -1.07$\pm$0.66 & -1.61$\pm$0.67 & 2.83$\pm$1.30 & 2.33$\pm$1.22 \\
		multiband coadd & N/A & -2.32$\pm$0.64 & -2.34$\pm$0.63 & 14.45$\pm$1.10 & 4.49$\pm$1.22\\

@@

\begin{figure*}
    %\hspace*{-3.5cm}
    \centering
	\includegraphics[width=\textwidth]{H158_meanshear_measured_properties_perbin_e1_v2.pdf}
    \caption{(Working on fixing the figure bounds)This shows the mean shear residual ($\Delta e_{1}$; $\langle e_{1,measured} \rangle - \langle e_{1,expected} \rangle$) as a function of various properties of galaxies for H158 single-band single-epochs, coadds, and multiband measurements. In this plot, the shear response is computed in each bin. The properties include the S/N, measured galaxy size $T_{gal,measured}$, input galaxy size $T_{gal,input}$, the shape and size of the input PSF.}
    \label{fig:meanshear}
\end{figure*}
